White Wines Quality Exploration by Sanjeev Yadav
========================================================

```{r packages, echo = F, message=F, warning=F}
knitr::opts_chunk$set(echo = F, message=F, warning=F)
# Load all of the packages that you end up using in your analysis in this code 
#chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. 
# This prevents the code from displaying in the knitted HTML output. You should
#set echo=FALSE for all code chunks in your file, unless it makes sense for your report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(dplyr)
library(gridExtra)
library(GGally)
library(RColorBrewer)
```

```{r Load_the_Data, echo=F}
# Load the Data
w <- read.csv("wineQualityWhites.csv")
# remove the index column
w$X <- NULL
```

This report explores a dataset containing quality and related chemical
properties for 4898 white wines.

# Univariate Plots Section

```{r dim, echo=F}
dim(w)
```

```{r structure, echo = F}
str(w)
```

```{r summary, echo = F}
summary(w)
```

Our dataset consists of 12 variables and 4898 observations. Most of the wines 
have quality 5.878, ranging from 3 to 9.  About 75% of the wines have residual
sugar below 10 gram/litre. Some wines have no citric acid in them. Maximum
alcohol content in white wines is 14.2%. 


```{r quality, echo=F, message=F, warning=F}
ggplot(w)+
  geom_bar(aes(factor(quality)))+
  xlab("quality")+
  theme_bw()
```

Quality distribution looks like a normal distribution with center at 6. Mode of
the distribution is also 6.

```{r pH1, echo=F, message=F, warning=F}
qplot(pH, data = w)+
  theme_bw()
```


```{r pH2, echo=F, message=F, warning=F}
ggplot(w)+
  geom_histogram(aes(pH),
                 binwidth = 0.01)+
  scale_x_continuous(breaks = seq(min(w$pH),
                                  max(w$pH),
                                  0.1))+
  theme_bw()
```

Distribution of pH is a normal distribution. Most of the wines are moderately
acidic with peaks around 3.12.

```{r sulphates1, echo = F, message=F, warning=F}
qplot(sulphates, data = w)+
  theme_bw()
```

```{r sulphates2, echo = F, message=F, warning=F}
ggplot(w)+
  geom_histogram(aes(sulphates),
                 binwidth = 0.01)+
  xlab("sulphates(g / dm^3)")+
  scale_x_continuous(breaks = seq(min(w$sulphates),
                                  max(w$sulphates),
                                  0.05))+
  theme_bw()
```

Distribution of sulphates is a right-skewed normal distribution, with most of
the wines having sulphates around 0.45 g/dm^3.

```{r density1, echo=F, message=F, warning=F}
qplot(density, data = w)+
  theme_bw()
```
```{r density2, echo=F, message=F, warning=F}
ggplot(w)+
  geom_histogram(aes(density),
                 binwidth = 0.0001)+
  xlim(min(w$density),1.005)+
  theme_bw()
```

```{r density.summary, echo = F, message=F, warning=F}
summary(w$density)
```

Density distribution is a normal distribution with mean = 0.994. There are some
outlier above 1.01.

```{r alcohol1, echo = F, message=F, warning=F}
qplot(alcohol, data = w)+
  theme_bw()
```


```{r alcohol2, echo=F, message=F, warning=F}
ggplot(w)+
  geom_histogram(aes(alcohol),
                 binwidth = 0.1)+
  scale_x_continuous(breaks = seq(min(w$alcohol),
                                  max(w$alcohol),
                                  0.5))+
  theme_bw()
```

Distribution of alcohol content is right skewed. High alcohol content is harmful
to health, this may be the reason why most of the wines have alcohol content
around 9.5.

```{r total.SO2.1, echo = F, message=F, warning=F}
qplot(total.sulfur.dioxide, data = w)+
  theme_bw()
```

```{r total.SO2.2, echo = F, message=F, warning=F}
ggplot(w)+
  geom_histogram(aes(total.sulfur.dioxide),
                 binwidth = 2)+
  scale_x_continuous(breaks = seq(0,250,25),
                     limits = c(0,250))+
  theme_bw()
```

```{r total.so2.summary, echo = F, message=F, warning=F}
summary(w$total.sulfur.dioxide)
```

The total sulfur dioxide follow a normal distribution with mean at 138.4. There
are some outliers above 250.

```{r free.SO2.1}
qplot(free.sulfur.dioxide, data = w)+
  theme_bw()
```
```{r free.SO2.2}
ggplot(w)+
  geom_histogram(aes(free.sulfur.dioxide),
                 binwidth = 2)+
  scale_x_continuous(breaks = seq(min(w$free.sulfur.dioxide),
                                  100,
                                  10),
                     limits = c(min(w$free.sulfur.dioxide),
                                100))+
  theme_bw()
```

```{r free.so2.summary}
summary(w$free.sulfur.dioxide)
```

The free sulfur dioxide has a normal distribution with mean around 35.3 and some
outlier above 100.

```{r chlorides1}
qplot(chlorides, data = w)+
  theme_bw()
```

```{r chlorides2, echo=F, message=F, warning=F}
ggplot(w)+
  geom_histogram(aes(chlorides),
                 binwidth = 0.001)+
  scale_x_continuous(breaks = seq(min(w$chlorides),
                                  0.2,
                                  0.02),
                     limits = c(min(w$chlorides),
                                0.2))+
  xlab("chlorides(g / dm^3")+
  theme_bw()
```

```{r chlorides.summary}
summary(w$chlorides)
```

50% of the wines have chlorides between 0.036 and 0.05. The distribution is
highly right skewed.

```{r res.sug1}
qplot(residual.sugar, data = w)+
  theme_bw()
```

```{r res.sug2, echo=F, message=F, warning=F}
ggplot(w)+
  geom_histogram(aes(residual.sugar),
                 binwidth = 0.02)+
  #xlim(min(w$residual.sugar),20)+
  scale_x_log10(breaks = c(1,2,5,10,20))+
  theme_bw()+
  xlab("residual sugar( g / dm^3)")
```

Using log scale to better understand the long tailed data. The transformed
distribution of residual sugar looks bimodal with peaks around 1.5 and around 10.

```{r citric.acid1}
qplot(citric.acid, data = w)+
  theme_bw()
```

```{r citric.acid2}
ggplot(w)+
  geom_histogram(aes(citric.acid),
                 binwidth = 0.01)+
  xlim(min(w$citric.acid),0.75)+
  xlab("citric acid (in g / dm^3)")+
  theme_bw()
```

```{r citric.acid.summary}
summary(w$citric.acid)
```

Citric acid follows a normal distribution with mean near 0.33 and some outliers
above 0.75. Using binwidth = 0.01 shows that there is an unexpected peak around
0.5.

```{r volatile.acidity1}
qplot(volatile.acidity, data = w)+
  theme_bw()
```

```{r volatile.acidity2}
ggplot(w)+
  geom_histogram(aes(volatile.acidity),
                 binwidth = 0.009)+
  scale_x_continuous(breaks = seq(min(w$volatile.acidity),
                                  max(w$volatile.acidity),
                                  0.1),
                     limits = c(0.08,0.68))+
  xlab("volatile acidity(acetic acid in g / dm^3)")+
  theme_bw()
```

Volatile acidity follows a right-skewed normal distribution with with peaks 
around 0.28.

```{r fix.acidity1}
qplot(fixed.acidity, data = w)+
  theme_bw()
```

```{r fixed.acidity2, echo=F, message=F, warning=F}
ggplot(w)+
  geom_histogram(aes(fixed.acidity),
                 binwidth = 0.1)+
  scale_x_continuous(breaks = seq(4.8,9.8,1),
                     limits = c(4.8,9.8))+
  xlab("fixed acidity (tartaric acid in g / dm^3)")+
  theme_bw()
```

The fixed acidity has normal distribution with mean near 6.8 and some outliers
above 9.8.

A new feature can be created by summing up all the acidities. The total acidity
is the amount of fixed acidity plus the volatile acidity. Fixed acidity includes
only tartaric acid and volatile acidity includes acetic acid. So, the total
acidity is the sum of all the acidities.

```{r total.acidity, echo=T}
w$total.acidity <- w$fixed.acidity + w$volatile.acidity + w$citric.acid
```

```{r total.acidity2}
qplot(total.acidity, data = w)+
  theme_bw()
```

```{r}
ggplot(w)+
  geom_histogram(aes(total.acidity),
                 binwidth = 0.01)+
  scale_x_continuous(breaks = seq(4,
                               15,
                               1))+
  theme_bw()
```
```{r}
summary(w$total.acidity)
```

It will be helpful to see the relationship between total acidity and quality of
wine. 

# Univariate Analysis

### What is the structure of your dataset?

There are 4898 white wines with 12 features(fixed acidity, volatile acidity,
citric acid, residual sugar, chlorides, free sulfur dioxide, total sulfur
dioxide, density, pH, sulphates, alcohol and quality). At least 3 wine experts
rated the quality of each wine, providing a rating between 0(very bad) and
10(very excellent). All the 11 chemical properties are numerical values of
different ranges.

It will be helpful to treat quality variable as ordered categorical variable.

```{r quality.level, echo=T}
w$quality.level <- ordered(w$quality, levels = c(3:9))
```


Other observations:

Most of the variables have normal distribution with some outliers.

### What is/are the main feature(s) of interest in your dataset?

The feature which we are trying to study is the quality of wine.Most of the
wines have average quality rating. We will use the given chemical properties
to predict the quality of wine. 

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

A good quality wine depends on the following factors: acid content and combined
(sugar + alcohol) content. Other interesting features for this investigation can
be: fixed.acidity, volatile.acidity, citric.acid, alcohol, residual.sugar and
quality.

### Did you create any new variables from existing variables in the dataset?

Total acidity is the sum of the acids in this dataset. It included volatile
acidity(acetic acid), fixed acidity(tartaric acid) and citric acid.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Residual sugar has the most unusual distribution. Using log10 scale on data
shows that residual sugar had a bimodal distrrbution. Most of the other
variables followed normal distribution, some of them were right skewed.

# Bivariate Plots Section

```{r ggpairs1, fig.height=12, fig.width=12}
set.seed(101)
ggpairs(w[sample.int(nrow(w), 1000),])
```
 

Density is strongly correlated with alcohol and residual sugar. This is an
unexpected correlation. Lets see correlation matrix for only these features.

```{r}
cor(w[,c("residual.sugar", "density", "alcohol", "quality", "total.acidity")])
```

```{r ggpairs2}
set.seed(101)
ggpairs(w[sample.int(nrow(w), 1000), c("residual.sugar", "density", "alcohol", "quality", "total.acidity")])
```

Lets see how these features are related with one another.

```{r alcohol.density}
ggplot(aes(alcohol,density), data = w)+
  geom_point(color = 'blue', alpha = 0.2, position = 'jitter')+
  coord_cartesian(xlim = c(quantile(w$alcohol, 0.05), quantile(w$alcohol, 0.95)),
                  ylim = c(quantile(w$density, 0.01), quantile(w$density, 0.99)))+
  geom_smooth(method = 'lm', color = 'red')+
  theme_bw()+
  ggtitle("Density vs Alcohol")
```

We can see a very strong correlation of -0.81 between the density and the
alcohol content.

```{r sugar.density}
ggplot(aes(residual.sugar, density),data = w)+
  geom_point(color = "blue",
             alpha = 0.1,
             position = 'jitter'
             )+
  coord_cartesian(xlim = c(quantile(w$residual.sugar, 0.01),quantile(w$residual.sugar, 0.99)),
                  ylim = c(quantile(w$density, 0.03),quantile(w$density, 0.97)))+
  geom_smooth(method = 'lm', color = "red")+
  theme_bw()+
  ggtitle("Density vs Residual Sugar")
```

There is a very strong correlation of 0.83 between the residual sugar and the
density of the wine.

```{r alcohol.quality}
ggplot(aes(alcohol, quality), data = w)+
  geom_point(color = 'blue', alpha = 0.2, size = 3, position = 'jitter')+
  coord_cartesian(xlim = c(quantile(w$alcohol, 0.05),
                           quantile(w$alcohol, 0.95)))+
  geom_smooth(method = 'lm', color = 'red')+
  theme_bw()+
  ggtitle("Quality vs Alcohol")
```

Very small but positive correlation can be seen between quality and alcohol
content. Quality increases gradually as alcohol content increases.

```{r density.quality}
ggplot(aes(density, quality), data = w)+
  geom_point(color = 'blue', alpha = 0.4, position = 'jitter')+
  coord_cartesian(xlim = c(quantile(w$density, 0.01),
                           quantile(w$density, 0.99)))+
  geom_smooth(method = 'lm', color = 'red')+
  theme_bw()+
  ggtitle("Quality vs Density")
```

There is a weak but negative correlation between the quality and the density
of the wine.

```{r acidity.quality}
ggplot(aes(total.acidity, quality), data = w)+
  geom_point(color = 'blue', alpha = 0.4, position = 'jitter')+
  coord_cartesian(xlim = c(quantile(w$total.acidity, .05), quantile(w$total.acidity, 0.95)))+
  geom_smooth(method = 'lm', color = 'red')+
  theme_bw()+
  ggtitle("Quality vs Total Acidity")
```

There is a weak correlation of -0.13 between quality and total acidity. The
almost horizontal line shows that there is a weak dependence of total acidity on quality.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

A very strong correlation was observed in case of alcohol. It had a strong
correlation with density and residual sugar. But there was no feature which had
a strong enough correlation with quality. It was difficult to assign a higher
significance level to a particular feature. 

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

The density had some interesting correlation with other features. Excluding the
main features, the density had the highest values for correlation co-efficients.

### What was the strongest relationship you found?

The strongest correlation was between density and residual sugar. Correlation
coefficient for this pair was 0.83. Density and alcohol also have high
correlation of -0.81.

# Multivariate Plots Section

```{r res.sug.density.quality}
ggplot(aes(density, residual.sugar, color = quality.level), data = w)+
  geom_point(alpha = 0.4, size = 1, position = 'jitter')+
  coord_cartesian(xlim = c(quantile(w$density, 0.05),
                           quantile(w$density, 0.95)))+
  geom_smooth(method = 'lm', color = 'red')+
  facet_wrap(~quality.level)+
  scale_color_brewer(palette = 'Set3')+
  theme_dark()+
  ggtitle("Residual Sugar vs Density for each Quality Level")
```

There is positive correlation between residual sugar and density for every
quality rating.

```{r alcohol.density.quality}
ggplot(aes(density, alcohol, color = quality.level), data = w)+
  geom_point(alpha = 0.5, size = 1, position = 'jitter')+
  coord_cartesian(xlim = c(quantile(w$density, 0.05),
                           quantile(w$density, 0.95)),
                  ylim = c(min(w$alcohol),max(w$alcohol)))+
  geom_smooth(method = 'lm', color = 'red')+
  facet_wrap(~quality.level)+
  scale_color_brewer(palette = 'Set3')+
  theme_dark()+
  ggtitle("Alcohol vs Density for each Quality Level")
```

There are very less wines with high quality and they all fall in the top left
region of the first quadrant here.

Now lets see the behaviour of quality vs the other features of interest, i.e.,
alcohol, residual sugar and total acidity.

```{r quality.alcohol.res.sug}
ggplot(aes(alcohol, quality, color = log10(residual.sugar)), data = w)+
  geom_point(position = 'jitter')+
  coord_cartesian(xlim = c(quantile(w$alcohol, 0.05),
                           quantile(w$alcohol, 0.95)))+
  scale_color_distiller(palette = 'Set4')+
  theme_bw()+
  ggtitle("Quality vs Alcohol by Log10(Residual Sugar)")
```

It looks like high quality wines have high residual sugar(the light green 
points near quality 8). Coloring by residual sugar does not provide a clear
separation between high quality and low quality wines.

```{r quality.res.sug.alcohol}
ggplot(aes(residual.sugar, quality, color = alcohol), data = w)+
  geom_point(position = 'jitter')+
  coord_cartesian(xlim = c(quantile(w$residual.sugar, 0.05),
                           quantile(w$residual.sugar, 0.95)))+
  scale_color_distiller(palette = 'Set4')+
  theme_bw()+
  ggtitle("Quality vs Residual Sugar by Alcohol Content")
```

Clearly high quality wines have high alcohol content(in the top left part of
graph). Some low alcohol wines are also there but there are very few.

```{r quality.alcohol.res.sug.total.acidity}
ggplot(aes(alcohol + log10(residual.sugar), quality, color = total.acidity),
       data = w)+
  geom_point(position = 'jitter')+
  coord_cartesian(xlim = c(quantile(w$alcohol + log10(w$residual.sugar), 0.05),
                           quantile(w$alcohol + log10(w$residual.sugar), 0.95)))+
  scale_color_distiller(palette = 'Set4')+
  geom_smooth(method = 'lm', color = 'blue')+
  theme_bw()+
  ggtitle('Quality vs combined effect of (alcohol + log10(residual sugar)) by total acidity')
```

Total acidity does not provide a clear separation between type of wines but this
graph is showing a weak positive correlation when combined effect of alcohol and
residual sugar is taken into account.

```{r quality.total.acidity.alcohol.log10(res.sug)}
ggplot(aes(total.acidity, quality, color = alcohol + log10(residual.sugar)),
       data = w)+
  geom_point(position = 'jitter')+
  coord_cartesian(xlim = c(quantile(w$total.acidity, 0.05),
                           quantile(w$total.acidity, 0.95)))+
  scale_color_distiller(palette = 'Set4')+
  geom_smooth(method = 'lm', color = 'blue')+
  theme_bw()+
  ggtitle("Quality vs Total Acidity by alcohol + log10(residual sugar) combined")
```

Not a strong relation but quality tends to decrease gradually with acidity
content. The same result was obtained in the bivariate plots section.

These plots do not provide a strong idea of how to predict the quality of a
wine. Very few relationships exist and they all have a very weak correlation
when it comes to predicting quality directly. 

```{r linear.models, echo = T}
m1 <- lm(quality ~ alcohol, data = w)
m2 <- update(m1, ~ . + log10(residual.sugar))
m3 <- update(m2, ~ . + total.acidity)
summary(m3)
```

# Multivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

As seen in bivariate plots section, density has a strong correlation with
residual sugar and alcohol. No other feature showed this much strong interaction.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

In order to predict the quality of wines, I tried to fit a linear model to see
the combined effect of alcohol and residual sugar on the quality of wine. The
model does not provide any good predictions, the R-squared values is 0.21 and
there is no feature which is statistically significant(3 stars for all features
corresponding to low significance level).

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(w)+
  geom_histogram(aes(residual.sugar),
                 binwidth = 0.02)+
  scale_x_log10(breaks = c(1,2,5,10,20))+
  xlab("residual sugar( g / dm^3)")+
  ylab("Wines Count")+
  theme_bw()
```

### Description One

The distribution of residual sugar appears bimodal. Maybe wines are categorized
on the basis of sweetness of wine. But there is not a clear separation between
those categories. Residual sugar contains outliers also.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(residual.sugar, density),data = w)+
  geom_point(color = "blue",
             alpha = 0.1,
             position = 'jitter'
             )+
  scale_x_log10(breaks = c(1,2,5,10,20))+
  coord_cartesian(xlim = c(quantile(w$residual.sugar, 0.01),
                           quantile(w$residual.sugar, 0.99)),
                  ylim = c(quantile(w$density, 0.03),
                           quantile(w$density, 0.97)))+
  geom_smooth(method = 'lm', color = "red")+
  xlab("Residual Sugar (g / dm^3)")+
  ylab("Density (g / cm^3)")+
  ggtitle("Density vs Log10 Residual Sugar")
```

### Description Two

Residual sugar has the strongest relationship with other features. It has almost
linear relationship with density with a correlation coefficient of 0.83.

### Plot Three

```{r echo=FALSE, Plot_Three}
ggplot(aes(total.acidity, quality, color = alcohol + log10(residual.sugar)),
       data = w)+
  geom_point(position = 'jitter')+
  coord_cartesian(xlim = c(quantile(w$total.acidity, 0.05),
                           quantile(w$total.acidity, 0.95)))+
  scale_color_distiller(palette = 'Set4',
                        name = "Alcohol(% by volume) + \nLog10(Residual Sugar)")+
  geom_smooth(method = 'lm', color = 'blue')+
  xlab("Total acidity (g / dm^3)")+
  ylab("Quality")+
  ggtitle("Quality vs Total Acidity by combined Alcohol Content + Log10 Residual Sugar")
```


### Description Three

Quality rating has a weak correlation with total acidity. High quality wines
have less acidity with high alcohol content and residual sugar value.

```{r}
m1 <- lm(quality ~ total.acidity + log10(residual.sugar) + alcohol,
         data = w)
summary(m1)
```

Linear model also provides a R-squared values of 21.1 %. Only 21% of the
variance can be explained by these predictors.

------

# Reflection

This dataset contains information about chemical properties of 4898 white wines.
Exploratory data analysis was carried to study the relationship between
features. Going through the internet helped to figure out how these features can
be use collectively to explain the quality of wines. Density has a strong
correlation with residual sugar and alcohol, but that did not help in predicting
the quality of wine. No feature could explain a good amount of variance in
quality. Finally the combined features of total acidity, residual sugar and
alcohol accounted for only 21% of the variance.

One limitation of this dataset is that the quality rating were more around
average rating. Most of the wines have rating of 6 and it was difficult to
generalize a linear model to predict the quality.

One addition which can be done in future analysis is the addition of new
features which better explain the variance in quality. Including more wines of
rating other than 6 will help to generalize a regression model trying to predict
the quality of wine.

# References

1. https://www.everwonderwine.com/blog/2017/1/14/is-there-a-relationship-between-a-wines-alcohol-level-and-its-sweetness?format=amp

2. http://winefolly.com/review/understanding-acidity-in-wine/